Dynamic Space Efficient Cuckoo Table -- DySECT when doubling is too much!

### 2017-04-17 paper/paper.tex
Laptop - paper changes +bibliography

@@

\begin{abstract}
Research on hash table data-structures has concentrated on optimizing
the speed and space efficiency for a long time.  Modern open
addressing techniques can fill a given table to more than 95\% before
the performance deteriorates significantly.  But once the threshold is
met, copying and migrating the whole table still seems to be the most
efficient answer.

To achieve any kind of efficiency, each such migration must be
amortized by newly inserted elements.  Therefore, each migration has
to increase the capacity by a significant amount -- usually, this
means doubling the capacity.  Growing by a large factor $\gamma$ leads
to bad space efficiency because the newly grown table is only filled
less than a factor of $\gamma^{-1}$.

The space efficiency is even worse during the migration, where
oftentimes both the new and the old table coexist at the same time
causing a massive space overhead.  To solve both of these issues, we
propose the a new hash table architecture that supports dynamic
growing while keeping a strict size constraint $\alpha \cdot n$
throughout its lifetime.  Even under these conditions, we can
guarantee constant lookup times, and amortized insertion times,
independent of $n$.
 \end{abstract}

@@

\section{Motivation}
\label{sec:motication}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similar tight bounds.

@@

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements.  Dynamic growing can be a
solution even in these instances.  It is completely independent from
the problem's domain and thus also works in cases where estimating the
number of elements can be hard.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and mgration.

\section{Related Work}
As previously stated, there has been a lot of research on hash tables,
since they were introduced in the 1950s.

@@

\subsection{Cuckoo Hashing}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables. The article TODO by TODO
(probably mitzenmacher with or without dietzfelbinger) et al. gives a
good overview over many cuckoo hashing variants.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to find $H$ independent possible positions. Each
element is stored in one of its positions.  Even if all possible
positions are occupied one can often move elements to create space for
the current element. We call this process displacing elements.

@@

The cost of growing the subtable is amortized by all insertions that
happened since the last subtable migration.  There are at least
$\alpha^{-1} \cdot s = \Omega(s)$ insertions between two migrations.
One migration takes $\Theta(s)$ time.  Appart from being amortized,
the migration is efficient in practise because it accesses cells in a
linear fashion making it really cache efficient.  Even in the target
table cells are accessed linearly.  Because of the way we assign
elements to buckets, one original bucket is split into two target
buckets.  Therefore, no displacements are necessary, since no bucket
can be overfilled.

@@

Another option, which offers some benefits is to use virtual memory,
and memory overcommiting.  The idea is that the operating system will
usually allow larger memory allocations than the machine has memory,
with the anticipation that not all the allocated memory will actually
be used.  Only memory pages, that are actually used will be mapped
from virtual to physical memory pages.  Thus from a strict reasource
point of view, the memory is not yet used.  Therefore, we can also
implement the subtables as sections within one huge allocation.  This
has the advantage, that the offset for each table can be quickly
computed, without looking it up from a table.

@@

\subsubsection*{Imbalance through Inhomogeneous Table Resolution}
\label{sec:inhom_res}
By growing subtables individualy we introduce a size imbalance
between subtables.  Large tables contain more buckets, but the number
of elements hashed to the table itself is not dependent on its size,
therefore, it is difficult to spread elements evenly among buckets.
Uneven bucket fill ratios can lead to longer insertion times.

@@

\subsubsection*{Imbalance through Size Changes}
In addition to the problem of inhomogenous tables, there is an
inherent balancing problem introduced by resizing subtables. It is
clear, that a newly grown table is not filled as densely as other
tables.  Since we double the table size, grown tables can only be
filled to about 50\%.

@@

DySECTs performance degrades earlier than cuckoo hashing.  One reason
for this is that our hash table consists of $T$ subtables with power
of two sizes and thus does not allow exactly $20M$ cells.  Therefore,
our table is slightly more full.  Another reason is that to get close
to $20M$ cells, we have TODO ??? large subtables (twice the size of
the others).  This introduces inhomogeneous table resolution (see
\ref{sec:inhom_res}), which decreases our performance compared to
cuckoo hashing.

Figure~\ref{fig:eps_find} shows that linear probing variants have bad
performance on find operations on highly filled tables.  It also shows
that unsuccessful find operations are often very slow when using
linear probing.  This is where the acceleration data-structure used by
hopscotch tables can significantly improve performance.  Both of these
effects, do not affect cuckoo variants like DySECT, that have
guaranteed constant running times for all find operations --
independant of the fill degree.

@@

\subsection{Word Count a Practical use Case}
One of the most common use cases for hash tables -- dynamic hash
tables in particular -- is to unify data elements according to their
key.  Oftentimes it is unclear how many unique keys there are (how
large the table will get).  In these cases, only dynamic hash tables
can remain space efficient.  In the following test, we take the first
block of the common crawl data-set (TODO REFERENCE) and compute a word
count of the contained words.  The chosen block contains around 240M
words, with around 20M unique words. For the test, we hash each word
and insert it's hash together with a counter.  Subsequent insertions
of a reocurring word increase this counter.  Similar to the growing
benchmark, we start with an empty table initialized for 50\,000
elements.

@@

\subsection{Conclusion}
\subsubsection*{Future Work}
Our Experiments clearly show, that practical running times can be
reached while growing dynamically.  But we were unable, to prove tight
theoretical performance bounds.

Even in practice, there remains a lot of open design space.  One could
use a similar aproach to dynamically grow different cuckoo hashing
variants e.g. concurrent cuckoo tables similar to Li et
al.~\cite{AlgorithmicImprovementsForFastConcurrentCuckooHashing} or
Nguyen and Tsigas~\cite{LockFreeCuckooHashing}; or a de-amortized
table proposed by Kirsch and
Mitzenmacher~\cite{UsingAQueueToDeamortizeCuckooHashingInHardware},
and refined by Arbitman et
al.~\cite{DeAmortizedCuckooHashingProvableWorstCasePerformanceAndExperimentalResults}.

@@

\bibliography{lipics-v2016-sample-article}

>>>>

% -*- mode: latex; mode:writegood mode: flyspell; ispell-local-dictionary: "en_US"; coding: utf-8 -*-

@@

\begin{abstract}
Research on hash table data-structures has concentrated on optimizing
the speed and space efficiency for a long time.  Modern open
addressing techniques can fill a given table to more than 95\% before
the performance deteriorates significantly.  But once the threshold is
met, copying and migrating the whole table still seems to be the most
efficient answer.
Growing by a constant factor $\gamma$ often leads
to bad space efficiency because the newly grown table is only filled
less than a factor of $\gamma^{-1}$.

The space efficiency is even worse during the migration, where
oftentimes both the new and the old table coexist at the same time
causing a massive space overhead.  To solve both of these issues, we
propose the a new hash table architecture that supports dynamic
growing while keeping a strict size constraint $\alpha \cdot n$
throughout its lifetime.  Even under these conditions, we can
guarantee constant lookup times, and amortized insertion times,
independent of $n$.

We achieve this goal by splitting the hash table into subtables that
grow independent from each other.  The cuckoo load balancing
algorithm can move elements from one table to the other.  This makes
sure that growing one subtable increases insertion opportunities even
for elements that are not hashed into this one table.
 \end{abstract}

@@

\section{Motivation}
\label{sec:motivation}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similar tight bounds.

@@

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements.  Dynamic growing can be a
solution even in these instances.  It is completely independent from
the problem's domain and thus also works in cases where estimating the
number of elements can be hard.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and migration.

\section{Related Work}
The usage of hash tables, and other hashing based algorithms has a
long history in computer science.  The classical methods results can
be found in all major algorithm textbooks (TODO ...).

Over the last one and a half decades, the field has regained
attention, both from theoretical and the practical point of view.  The
initial point of innovation was the power of two
choices~\cite{ThePowerOfTwoChoicesInRandomizedLoadBalancing}.
This concept led to the development of cuckoo hashing by Pagh and
Rodler~\cite{CuckooHashing}.  This technique reopened research into
more space efficient hash tables.  It proofed theoretically
interesting, because probabilistic bounds for the maximum fill degree
are highly non-trivial TODO SOME BOUNDS PAPERS.

Cuckoo hashing spawned many variants, that further improve its
performance.  The most notable one is probably bucket cuckoo hashing
($d$-ary cuckoo hashing), a variant that groups cells together in
buckets thus improving cache usage, and increasing displacement
opportunities TODO BUNCH OF PAPERS.  This further improves the maximum
fill degree. Nowadays, bucket cuckoo hashing is by far the most common
variant. There are even further variations on bucket cuckoo hashing,
like the generalization to continuous buckets.

Further adaptations to cuckoo hashing include: a de-amortization
technique that allows a provable worst case analysis of
insertions~\cite{UsingAQueueToDeAmortizeCuckooHashingInHardware,
  DeAmortizedCuckooHashingProvableWorstCasePerformanceAndExperimentalResults};
and multiple concurrent implementations either powered by bucket
locking, transactional
memory~\cite{AlgorithmicImprovementsForFastConcurrentCuckooHashing},
or fully lock-less~\cite{LockFreeCuckooHashing}.

Some non-cuckoo space efficient hash tables continue to use linear
probing variants.  Robin Hood hashing is a technique that was
originally introduced in 1985 by Celis et
al.~\cite{RobinHoodHashing}. It has regained some popularity in
recent years -- mainly for its interesting theoretical
properties and the possibility to reduce the inherent variance of
linear probing.  Hopscotch~\cite{HopscotchHashing} hashing is a
technique, that originated in the area of concurrent hash tables, but
it can also lead to some densely filled hash tables.  For this reason
it is often used as a technique to achieve space efficient hash
tables.  This is only possible in a small number of circumstances, as
the stored bitmaps used in hopscotch hashing produce a significant
memory overhead.

All these publications show that there is a clear interest in the
development of space efficient hash tables.  Dynamic hash tables on the
other hand seem to be considered a solved problem.  There is an
important paper by Dietzfelbinger at
al.~\cite{DynamicPerfectHashingUpperAndLowerBounds} that takes on the
problem of dynamic hash tables from a theoretical perspective.  The
paper predates cuckoo hashing, and much of the attention for space
efficient hashing. Thus all memory bounds presented are given without
tight constant factors.  This lack of development is where we pick up
and offer fast solutions that support dynamic growing with tight space
bounds.

@@

\subsection{Cuckoo Hashing}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to find $H$ independent possible positions. Each
element is stored in one of its positions.  Even if all possible
positions are occupied one can often move elements to create space for
the current element. We call this process displacing elements.

@@

The cost of growing the subtable is amortized by all insertions that
happened since the last subtable migration.  There are at least
$\alpha^{-1} \cdot s = \Omega(s)$ insertions between two migrations.
One migration takes $\Theta(s)$ time.  Apart from being amortized,
the migration is efficient in practice because it accesses cells in a
linear fashion making it really cache efficient.  Even in the target
table cells are accessed linearly.  Because of the way we assign
elements to buckets, one original bucket is split into two target
buckets.  Therefore, no displacements are necessary, since no bucket
can be overfilled.

@@

Another option, which offers some benefits is to use virtual memory,
and memory overcommiting.  The idea is that the operating system will
usually allow larger memory allocations than the machine has memory,
with the anticipation that not all the allocated memory will actually
be used.  Only memory pages, that are actually used will be mapped
from virtual to physical memory pages.  Thus from a strict resource
point of view, the memory is not yet used.  Therefore, we can also
implement the subtables as sections within one huge allocation.  This
has the advantage, that the offset for each table can be quickly
computed, without looking it up from a table.

@@

\subsubsection*{Imbalance through Inhomogeneous Table Resolution}
\label{sec:inhom_res}
By growing subtables individually we introduce a size imbalance
between subtables.  Large tables contain more buckets, but the number
of elements hashed to the table itself is not dependent on its size,
therefore, it is difficult to spread elements evenly among buckets.
Uneven bucket fill ratios can lead to longer insertion times.

@@

\subsubsection*{Imbalance through Size Changes}
In addition to the problem of inhomogeneous tables, there is an
inherent balancing problem introduced by resizing subtables. It is
clear, that a newly grown table is not filled as densely as other
tables.  Since we double the table size, grown tables can only be
filled to about 50\%.

@@

DySECT's performance degrades earlier than cuckoo hashing.  One reason
for this is that our hash table consists of $T$ subtables with power
of two sizes and thus does not allow exactly $20M$ cells.  Therefore,
our table is slightly more full.  Another reason is that to get close
to $20M$ cells, we have TODO ??? large subtables (twice the size of
the others).  This introduces inhomogeneous table resolution (see
\ref{sec:inhom_res}), which decreases our performance compared to
cuckoo hashing.

Figure~\ref{fig:eps_find} shows that linear probing variants have bad
performance on find operations on highly filled tables.  It also shows
that unsuccessful find operations are often very slow when using
linear probing.  This is where the acceleration data-structure used by
hopscotch tables can significantly improve performance.  Both of these
effects, do not affect cuckoo variants like DySECT, that have
guaranteed constant running times for all find operations --
independent of the fill degree.

@@

\subsection{Word Count a Practical use Case}
One of the most common use cases for hash tables -- dynamic hash
tables in particular -- is to unify data elements according to their
key.  Oftentimes it is unclear how many unique keys there are (how
large the table will get).  In these cases, only dynamic hash tables
can remain space efficient.  In the following test, we take the first
block of the common crawl data-set (TODO REFERENCE) and compute a word
count of the contained words.  The chosen block contains around 240M
words, with around 20M unique words. For the test, we hash each word
and insert it's hash together with a counter.  Subsequent insertions
of a re-occurring word increase this counter.  Similar to the growing
benchmark, we start with an empty table initialized for 50\,000
elements.

@@

\subsection{Conclusion}
Our Experiments have shown that our DySECT hash table is by far the
best performing hash table on all instances that make use of its
dynamic capabilities.  It performs up to TODO $\%$ better, than the
next best performing table.

.........

But there remains much to do. We were unable, to prove tight
theoretical performance and density bounds.

Even in practice, there remains a lot of open design space.  One could
use a similar approach to dynamically grow different cuckoo hashing
variants e.g. concurrent cuckoo tables similar to Li et
al.~\cite{AlgorithmicImprovementsForFastConcurrentCuckooHashing} or
Nguyen and Tsigas~\cite{LockFreeCuckooHashing}; or a de-amortized
table proposed by Kirsch and
Mitzenmacher~\cite{UsingAQueueToDeAmortizeCuckooHashingInHardware},
and refined by Arbitman et
al.~\cite{DeAmortizedCuckooHashingProvableWorstCasePerformanceAndExperimentalResults}.

............

@@

\bibliography{paper}
